elsevier_MANUSCRIPT

### 2025-03-20 elsevier_MANUSCRIPT.qmd
Update elsevier_MANUSCRIPT.qmd

---
title: "elsevier_MANUSCRIPT"
format: html
---

>>>>

---
title: "elsevier_MANUSCRIPT"
format:
  pdf: default
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Journal of Hydrology
      formatting: preprint
      model: 1p
      layout: onecolumn
      cite-style: number
      highlights:
        - Approximately 32% of Arizona's streamflow originates from groundwater discharge, with significant spatial and climatic variability. 
        - A machine learning model (XGBoost) successfully predicted long-term BFI in ungauged basins, addressing gaps in a sparse streamgage network. 
        - Base-flow trends align most closely with precipitation changes, emphasizing the role of climate in shaping groundwater-surface water dynamics in dryland regions.
---

## Introduction {#sec-intro}

Dryland regions, encompassing arid, semi-arid, hyper-arid, and dry sub-humid systems, account for 40% of the Earth’s land surface. These regions are home to approximately 2 billion people globally and constitute the largest terrestrial biome [@iucn-drylands-2019]. Despite supporting diverse ecosystems and human populations, dryland regions face mounting hydrologic challenges exacerbated by increasing urbanization, expanding agricultural activities, and climate-induced amplification of precipitation patterns [@taylor2013]. This water scarcity is intensifying due to the compounding effects of climate variability and increased groundwater extraction [@taylor2013]. In drylands, groundwater serves as a vital resource for sustaining ecosystems and meeting human needs [@scanlon2006; @yao2018].

Base flow is the sustained portion of streamflow in the absence of runoff that is derived from groundwater discharge [@usgs-glossary-2018]. Base flow is critical to maintaining seasonal low-flow regimes, supporting aquatic ecosystems, and facilitating the transport of nutrients and chemicals. Base-flow contribution to streamflow can be highly variable spatially [@singh2018; @bosch2017; @beck2013], and temporally [@ficklin2016; @tan2020]. Increasing groundwater extraction, changes in land cover/land use, and changes in precipitation patterns due to climate change affect the timing and volumes of base flow [@tan2020; @taylor2013]. Effective management of water quantity and quality depends on understanding seasonal and interannual base-flow patterns and long-term changes in base-flow behavior.

The Base-Flow Index (BFI) is the ratio of the long-term mean base-flow volume to the long-term total streamflow volume expressed as a percentage. BFI serves as a normalized measure of groundwater contribution interannually or between basins. BFI is determined by hydrograph separation and is influenced by the climate and physiographic characteristics of a catchment [@neff2005; @beck2013; @singh2018]. Between catchments, base flow fluctuates according to changes in the moisture content of the vadose zone, influenced by varying levels of evapotranspiration and aquifer storage dynamics [@bosch2017]. Since BFI calculations rely on instrumented stream records, it remains unknown for ungauged catchments, which encompass most of the earth's land surface [@fekete2007]. Addressing this information gap is integral to approaching a comprehensive understanding of groundwater dynamics globally.

Advancements in machine learning provide tools to predict hydrologic indices in ungauged basins, addressing the limitations of sparse streamgage networks. To tackle the challenge of quantifying base flow in ungauged catchments, numerous studies have applied both regression and machine learning methods. @ahiablame2013 found that using a regression model to estimate annual base flow of ungauged catchments was reasonably easy and accurate. @beck2013 overcame the nonlinearity of basin characteristics and improved results of multivariate analyses by using artificial neural networks to estimate BFI globally. @singh2018 implemented a random forest algorithm to predict long-term BFI for ungauged catchments across New Zealand. These applications demonstrate the versatility and effectiveness of machine learning in capturing complex ecohydrologic dynamics and improving our understanding of groundwater contributions to streamflow.

This study develops a technique for estimating BFI in ungauged basins across Arizona and evaluates the state’s long-term BFI. Regional trends in base flow and BFI at instrumented sites are also analyzed, with these trends being linked to coincident trends in precipitation, reference evapotranspiration (ET~O~), and temperature over the same periods. Using a machine learning model trained on the hydrogeologic characteristics of surface water basins, we estimate the mean BFI for ungauged basins from 1991 to 2020. This approach helps address the spatial gaps in the streamgage network, which is relatively sparse across the state. The results offer novel insights into low-flow processes in both gauged and ungauged basins, enhancing our understanding of climate controls on consistent flows in Arizona. This study contributes to a more comprehensive view of hydrological dynamics in the context of arid and semi-arid landscapes.

## Methods {#sec-methods}

### Study Area {#sec-study-area}

The state of Arizona, located in the southwest United States, covers a total area of 295,253 km$^2$. Arizona is divided into two primary physiographic provinces: the Colorado Plateau in the northeast, and the Basin-and-Range region in the west and south. The Central Highlands is a transition zone consisting of scattered basins separated by the mountainous foothills of the Mogollon Rim. The Colorado Plateau is dominated by high-elevation desert with an average elevation of 1,936 masl (6,352 ft). Mean temperatures range from -6$^\circ$C (20$^\circ$F) to 26$^\circ$C (80$^\circ$F) and it averages 580 mm (23 in) of precipitation. The Basin-and-Range region has a semi-arid to arid climate with an average elevation of 490 masl (1600 ft). Average temperature ranges from 15$^\circ$C (60$^\circ$F) to 43$^\circ$C (110$^\circ$F) and the region averages 200 mm (8 in) of precipitation annually [@az_climate2024].

![Map of Arizona and US Geological Survey (USGS) streamgages used in this study. 8-digit HUC subbasin boundaries and physiographic regions shown.](images/StudyArea_20241203.png){#fig-study-area}

Arizona's hydrology varies seasonally and spatially between its physiographic regions. In the summer, localized and intense convective precipitation events are driven by the North American Monsoon, while in the winter, orographic precipitation comes from Pacific frontal systems [@eastoe2019mtnblock]. While monsoonal precipitation can account for up to 50% of annual precipitation, evaporation and dry preceding soil properties leads to most precipitation becoming runoff [@sheppard2002]. As such, 94% of streams in Arizona are ephemeral or intermittent [@epa_AZephemeral]. Much of the hydrology of Arizona is snow-melt derived, driven by spring melt from the high-elevation Colorado Plateau winter snowpack. While winter precipitation provides only 30% of annual averages, it provides the majority of water for natural reservoirs [@sheppard2002].

### Data {#sec-data}

Daily observed streamflow data obtained from the United States Geological Survey (USGS) National Water Information System (NWIS) were used in this study. Streamgages were selected depending on criteria to ensure the applicability of each site. Following the findings of @odonnell2016, which determined that 8–10 years of calibration data are necessary to account for climate variability in paired watershed studies in the region, a minimum record length of 10 years was required. Additionally, years with more than 30 missing days of streamflow data were excluded from the analysis. This study focuses on natural, streamflow-influencing dynamics, so streamgages affected by regulation or diversions were excluded. The regulated river streamgages were identified through annual reports on water data published by the USGS [@usgs2010]. Furthermore, streamgages along the Colorado River were omitted because they represent managed flows governed by the Colorado River Compact. After applying these selection criteria, 205 USGS streamgages with acceptable periods of record were included in the study (@fig-study-area). Periods of record ranged from 10 to 112 years, with a median of 28 years.

While our data selection ensures a robust record for this analysis, it also highlights gaps in spatial coverage that the machine learning model can address. Arizona has 184 active USGS streamflow stations (as of 2024) across an area of 295,253 km², whereas Indiana (a humid state in the U.S.), for example, has 189 active stations within an area of 94,326 km². This translates to a streamgage density of approximately 2.004 gages per 1,000 km² in Indiana, over three times higher than Arizona's density of 0.623 gages per 1,000 km². This stark difference underscores the relative sparsity of Arizona's streamgage network, particularly in the context of its larger geographic area and the unique hydrological challenges posed by its arid and semi-arid landscapes. <!--# Add something here to discuss the application in broader semi-arid low streamgage regions globally -->

Watersheds across the United States are delineated by the USGS using a hydrologically-defined network. This system delineates the country using hierarchical hydrologic unit codes (HUCs), where each subsequent basin includes the digits of the enclosing basin. Here, 8-digit HUCs (HUC 8s) are used to divide Arizona into 84 sub-basins that are fully or partially in the state (@fig-study-area). These HUC 8 sub-basins are analogous to medium-sized river basins and are defined by surface water characteristics.

Annual precipitation and temperature data came from the PRISM climate group at Oregon State University at a resolution of 4 km (<https://prism.oregonstate.edu;> [@daly2008]). The PRISM dataset provides valuable insights into regional climate in ungauged regions and has been shown to perform well across the southwestern US [@buban_PRISM]. Instead of the water year, PRISM data uses a calendar-year format, which was adopted for consistency in the water balance. Although this may introduce challenges in the annual estimates due to inter-annual snow storage, the use of long-term annual averages is likely to reduce any potential errors [@reitz2017].

Annual reference evapotranspiration (ET~O~) data came from TerraClimate, a 4-km grid climatological data set [@abatzoglou2018]. TerraClimate uses a Penman-Monteith approach to generate a reference evapotranspiration. The ET~O~ values were calculated assuming a reference grass surface across the landscape with unlimited water. In the drylands of the southwestern US, ET~O~ typically exceeds precipitation annually [@zomer2022].

A 30-meter resolution Digital Elevation Model (DEM) of Arizona was used to derive key basin characteristics: basin area, average slope, and the proportion of each basin oriented toward north or south aspects. Various geospatial variables, such as aspect, were disaggregated then averaged to assess the areal percentage of each sub-variable within individual HUC 8 basins. By calculating these percentages, we aimed to get a more comprehensive understanding of landscape composition across space. Land cover was acquired from USGS-NLCD (National Land Cover Database), hydrologic soil group from SSURGO (Soil Survey Geographic Database), and underlying geology and karst from USGS were all similarly averaged across the basins. Aggregating variables to align with the HUC 8 boundaries allowed for more precise predictions of BFI by integrating spatial variations within each basin.

### Base-flow separation {#sec-bf_sep}

Directly measuring base flow and BFI presents unique challenges [@eckhardt2008]. The technique chosen to separate base flow has been shown to affect results, and the choice of base-flow separation method is subjective since ‘true’ BFI values are not known [@beck2013]. However, many methods have been developed to estimate these values. These methods include the use of tracers [@gonzales2009], graphical interpolation [@instituteofhydrology1980; @sloto1996], and digital filters [@arnold1995; @eckhardt2005; @lyne1979; @nathan1990]. These techniques have varying levels of applicability depending on the spatial scale, time span, and the scope of the study. Comparisons of various base-flow separation techniques have been made in previous studies [e.g. @eckhardt2005; @eckhardt2008; @nathan1990]; this study does not explore the superiority of different methods.

Base flow was calculated using a single-parameter, recursive digital filter technique from @nathan1990. This base-flow separation technique is based on a recursive digital filter used in signal analysis that separates high-frequency signals (quickflow) from low-frequency signals (base flow) [@lyne1979]. @eckhardt2023 noted that recursive digital filters lack a physical basis, but as the method is easy to automate, objective, and repeatable, it is appropriate for a regional-scale study. The Lyne-Hollick filter has been used in multiple studies [@ARNOLD200021; @santhi2008; @bloomfield2009; @singh2018], and it takes the form of

$$
b = \alpha b_{k-1} + \frac{1-\alpha}{2}(Q_k + Q_{k-1})
$$ {#eq-bf-separate}

where $b$ is base flow, $\alpha$ is the filter parameter, $Q$ is the total streamflow, and $k$ is the time step. A filter parameter $\alpha$ of 0.925 was used as in @nathan1990 and @fuka2014ecohydrology. The filter was run three times (forward, backward, forward) to attenuate the base-flow signal.

### Machine Learning

The implementation of machine learning models to predict hydrologic indices has been successful in past studies [@singh2018; @schmidt_ML_2020; @Rozos2021Machine]. In this work, we used the eXtreme Gradient Boosting (XGBoost) algorithm [@chen2016xgboost] to predict BFI at ungauged locations using catchment characteristics as predictors [@tbl-predictors]. The XGBoost algorithm is a decision tree-based ensemble algorithm, which can be adapted for either regression or classification problems. This algorithm iteratively builds an ensemble of decision trees, where each tree corrects errors from previous trees to improve predictions [@chen2016xgboost]. Its efficiency, scalability, and robustness have made it increasingly popular in recent years, with successful applications in environmental modeling tasks such as streamflow forecasting [@szczepanek2022daily; @ni2020streamflow] and land use/land cover classification [@Georganos2018Very].

XGBoost operates by leveraging gradient boosting on decision tree algorithms, combining multiple low-variance models to produce a robust overall prediction. Gradient boosting works iteratively: the initial tree is trained on the target values, while subsequent trees are trained on the residual errors of the preceding tree. Each tree is assigned a weight based on its contribution to reducing error, and these weights are used to determine the influence of each tree in the final model. The ultimate prediction is made by aggregating the outputs of all $n$ weighted trees in the ensemble. In this study, the trained XGBoost model is used to predict BFI in ungauged catchments based on geospatial and hydroclimate predictor variables (@tbl-predictors). Certain features were further subdivided according to their areal coverage within each basin (e.g. land cover was divided into 16 subdivisions). This approach allowed the model to capture finer-scale spatial variability and improve predictive accuracy.

|   | Variable | Source | Geoprocessing |
|----|----|----|----|
| Hydroclimate | Precipitation\* | PRISM | Basin average |
|  | Mean Temperature\* | PRISM | Basin average |
|  | Reference Evapotranspiration\* | TerraClimate | Basin average |
| Geospatial | Elevation\* | DEM | Basin average |
|  | Area | DEM | Basin average |
|  | Slope | DEM | Basin average |
|  | Aspect | DEM | Percent areal coverage |
|  | Land Cover\* | NLCD | Percent areal coverage |
|  | Hydrologic Soil Group\* | SSURGO | Percent areal coverage |
|  | Geology | USGS | Percent areal coverage |
|  | Karst | USGS | Percent areal coverage |

: Basin-characteristic variables used as initial features in XGBoost model. Starred features are maintained in the final, dimensionality-reduced model. {#tbl-predictors}

Our initial training dataset comprised 7,724 observations across 45 variables. To optimize the model's performance, we first conducted an exhaustive grid search combined with 5-fold cross-validation to identify the optimal hyperparameter values. The hyperparameters evaluated included the learning rate ($\eta$), minimum split loss ($\gamma$), maximum tree depth, minimum child weight, and the number of trees. While not an exhaustive list of all possible XGBoost hyperparameters, the range of values explored provided sufficient variation to ensure the selection of a high-performing model. The optimal hyperparameters were determined to be: 700 trees, a learning rate ($\eta$) of 0.05, a minimum split loss ($\gamma$) of 0.075, a maximum tree depth of 7, and a minimum child weight of 5. Using these values, the XGBoost model was trained on the dataset with 10-fold cross-validation.

$K$-fold cross-validation provides an unbiased estimate of a model's accuracy on unseen data, while also insuring against overfitting or underfitting. In this approach, the data are randomly divided into $k$ folds of equal size. The model is trained on $k-1$ folds and tested on the remaining fold, referred to as the validation set. This process is repeated $k$ times, with each fold serving as the validation set exactly once. Each iteration trains an independent model with the same hyperparameters but using a different subset of training data. By averaging the model performance across all $k$ folds, we achieve a robust and reliable estimate of predictive accuracy. Root mean squared error (RMSE) was used as the performance metric for both model optimization and evaluation. RMSE provided a consistent and interpretable measure of the model’s accuracy throughout the training and validation process.

![Schematic of $k$-fold cross validation. The dataset is randomly divided into $k$ stratified folds. Each fold serves as the validation set once, while the remaining folds are combined to create a training set for model development. Performance metrics for the test set are calculated and recorded, and this process is repeated for all $k$ folds.](images/k-fold_fig.png){#fig-k-fold fig-align="center" width="100%"}

#### Feature Selection {#sec-feat-selection}

Machine learning models are prone to overfitting, especially when provided with a large set of predictive features [@Ying_2019]. Overfitting can degrade model performance on unseen data and increase the demand for computational resources and memory storage [@feat_selec2017]. Dimensionality reduction offers a robust solution to these challenges and generally falls into two broad categories: feature extraction and feature selection.

Feature extraction involves transforming the original dataset into a lower-dimensional feature space. However, this process generates new features that often lack the physical interpretability of the original variables. In contrast, feature selection identifies a subset of the original features, preserving their physical meaning while improving model readability and interpretability [@feat_selec2017]. In this study, supervised feature selection was employed to reduce the number of predictors, which enhanced learning performance, reduced computational costs, and mitigated overfitting.

To begin, an initial model was trained using the full feature set of 45 predictors (@tbl-predictors). A feature selection method based on feature importance was then applied to identify and remove less relevant and noisy features. Feature importance scores quantify the contribution of individual features—either positively or negatively—to the model's predictions [@ML_interpretability2019]. In this analysis, SHAP (SHapley Additive exPlanations) values were used to compute feature importance scores [@SHAP_values].

SHAP values is a method to explain the prediction of an individual instance by calculating the contribution of each feature to that prediction. The method is based on coalition game theory and is discussed further in @SHAP_values. Here, SHAP values are used for global interpretation of feature importance and feature effects on the model. Global feature importance is produced by the absolute Shapley values of each feature across the dataset, providing a list of features in order of most to least important. Feature effects provide an indication of the relationship between the value of a predicting feature and its impact on the prediction.

The ten most important features [@fig-shap_values] were selected based on their SHAP values and used to train a subsequent model [@tbl-predictors]. This refined model demonstrated improved performance and reduced computational time compared to the initial model. The final model, trained on this optimized feature subset, was ultimately used for the analysis presented here.

### Statistical Analyses

Statistical analyses were conducted on annual BFI and base-flow values from instrumented streamgages to identify temporal trends using the Mann-Kendall nonparametric trend test [@kendall1970; @mann1945nonparametric]. This test detects monotonic trends in datasets that are non-parametric and assumes the absence of autocorrelation among observations. This test is widely used in studies of this nature [@ficklin2016; @ayers2019; @woodhouse_udall22].

To check for autocorrelation, we applied the Durbin-Watson test, which revealed significant autocorrelation at four streamgages on an annual basis. Of these, only one streamgage (09486500 - Santa Cruz River at Cortaro, AZ) showed a significant trend in BFI. This streamgage was excluded from the trend analysis, as autocorrelation could inflate the variance of the Mann-Kendall statistic, potentially leading to biased trend estimates [@MK_Hamed1998]. Trends with a $\rho \le 0.05$ are considered significant.

## Results

### BFI of Ungauged Catchments

#### Model Validation

Predicted values of BFI are plotted against observed values for the entire period of record of the instrumented dataset in @fig-actual_predicted . The agreement between "out-of-bag" predictions (blind cross-validation, treating each site as ungauged) and observed values is acceptable (R^2^ = 0.764) indicating that the model performs well across the full dataset. The overall RMSE is 0.129 and the overall percent bias (pbias) is -5.6%. Model performance metrics across various classifications are summarized in @tbl-performance. These performance metrics demonstrate that the regional model performs consistently well across different spatial and climatic classifications. However, the negative pbias values across all classifications, along with the overall pbias, indicate a systematic underprediction of BFI by the model. Categories with relatively lower R^2^ and Nash-Sutcliffe Efficiency (NSE) values also exhibit higher biases, reflecting weaker model performance in those specific contexts.

![Linear relationship between observed BFI and predicted BFI. The solid line is the 1:1 line, the dashed, green line is regressed to the data.](images/actual-predicted.png){#fig-actual_predicted}

| **Classification Group** | **n** | **R2** | **MSE** | **RMSE** | **MAE** | **NSE** | **pbias** |
|---------|---------|---------|---------|---------|---------|---------|---------|
| Climate - Monsoon Dominated | 3039 | 0.633 | 0.016 | 0.126 | 0.074 | 0.619 | -13.7 |
| Climate - Snowmelt Dominated | 4685 | 0.733 | 0.015 | 0.121 | 0.087 | 0.725 | -3.5 |
| PhysRegion - Basin&Range | 6147 | 0.733 | 0.016 | 0.127 | 0.084 | 0.724 | -6.3 |
| PhysRegion - CO Plateau | 1577 | 0.846 | 0.011 | 0.104 | 0.073 | 0.843 | -3.8 |
| Climate - Warm-Wet | 1506 | 0.693 | 0.014 | 0.117 | 0.077 | 0.685 | -8.2 |
| Climate - Warm-Dry | 2351 | 0.693 | 0.022 | 0.147 | 0.092 | 0.675 | -11.9 |
| Climate - Cool-Wet | 2350 | 0.738 | 0.011 | 0.106 | 0.078 | 0.736 | -1.7 |
| Climate - Cool-Dry | 1517 | 0.831 | 0.012 | 0.111 | 0.078 | 0.827 | -4.3 |
| Slope - High | 3795 | 0.776 | 0.012 | 0.111 | 0.079 | 0.771 | -3.3 |
| Slope - Low | 3929 | 0.724 | 0.018 | 0.133 | 0.085 | 0.713 | -9.1 |

: Performance of model predictions for BFI for all sites split by various classifications. n is number of observations, R^2^ is the coefficient of determination of a linear regression, MSE is mean-squared-error, RMSE is root-mean-squared-error, MAE is mean-absolute-error, NSE is Nash-Sutcliffe efficiency, and pbias is percent bias. {#tbl-performance}

#### Predictor Importance {#sec-predictor-importance}

The predictors used to estimate BFI at ungauged sites were evaluated for their importance in the final XGBoost model, as illustrated in @fig-shap_values. The most influential feature for predicting long-term BFI is basin elevation. While elevation itself does not directly affect base-flow characteristics, it has consistently been identified as a key predictor in previous BFI studies [@singh2018; @beck2013]. The importance of elevation aligns with findings from @beck2013, highlighting its role as a proxy for climate variables such as temperature, precipitation, and snowpack duration. Seasonal snowpack duration, in particular, has been shown to strongly correlate with springflow and groundwater recharge in this region [@donovan_karst_2022]. This relationship is further supported by hydroclimate features where higher temperatures tend to negatively influence BFI, while precipitation exhibits a mixed influence. In some cases, higher precipitation values correlate with lower BFI, likely due to a larger proportion of precipitation contributing to runoff rather than infiltration and groundwater recharge.

![SHAP value plot of features used in final model. Land cover features are indicated by the percentage of cover by each land cover type and soil types are defined by hydrologic soil group.](images/shap_summary.png){#fig-shap_values}

Land cover and land use predictors also play a significant role in BFI estimation. Analysis of SHAP values indicates that a higher percentage of evergreen forest positively influences BFI predictions, while higher proportions of shrubland and developed land exert a negative influence. Similarly, hydrologic soil types show distinct trends in their impact on BFI. Soil Type C, characterized by moderately high runoff potential (20-40% clay), tends to negatively influence BFI. In contrast, Soil Type A, which has low runoff potential and facilitates rapid water infiltration, exhibits a mixed influence [@nrcs2009]. The mixed effects of Soil Type A are likely due to SHAP values capturing interactions between features rather than direct relationships. For example, regions dominated by Soil Type A may also have steep slopes or sparse vegetation. Additionally, the relationship between Soil Type A and BFI is likely non-linear and influenced by the complex dynamics of regional base-flow processes.

#### Predicted Long-term BFI

![Predicted long-term BFI values for 8-digit HUC (1991-2020)](images/BFI_HUC8_20250129.png){#fig-bfi-huc}

The regionalized (HUC-8) long-term BFI (1991–2020) is shown in @fig-bfi-huc. Basins with high BFIs, such as those along the Grand Canyon in the northwestern part of the study area, indicate greater surface water and groundwater interaction. Elevated BFI values are also observed along portions of the Mogollon Rim, a heavily forested region with high precipitation that marks the transition between physiographic regions. Additionally, headwater regions of perennial rivers tend to exhibit higher BFI values. In contrast, low BFI values are found in areas like the Defiance Plateau in northeastern Arizona and the arid southern regions of the state. While BFI values and their spatial patterns may align with certain terrain features or climatic patterns, it is important to recognize that BFI is driven by complex interactions among multiple factors [@fig-shap_values].

### BFI of Gauged Catchments

![Long-term BFI for the period of record from instrumented stream flow data.](images/BFI_Instrumented_20241203.png){#fig-instrumented-bfi}

The long-term BFI for the 205 gauged reaches across Arizona is illustrated in @fig-instrumented-bfi . The long-term mean BFI is 0.32, indicating that \~32% of long-term streamflow in Arizona likely originates from groundwater discharge and other delayed sources. The highest BFI values (\>0.9) are found along the Grand Canyon in northwestern Arizona. The highly karstic geology of this region facilitates the rapid movement of subsurface flow to surface water and spring outlets [@chambless_deep-karst_2023]. Relatively high BFI values (\>0.8) are found at the spring-fed headwaters of the Verde River (Del Rio Spring) and the spring-fed headwaters of Fossil Creek. These results are consistent with interpolated BFI values reported by @wolock2003.

The stream reaches of the Little Colorado River Basin (northeastern Arizona) indicate consistently low BFI values (\< 0.2). This is likely due to low-yielding perched aquifers underlying the Defiance Plateau in northeastern Arizona, which are hydrologically connected to surface streams, while the high-yield, confined regional aquifer is much deeper [@blanchard02]. A notable tendency emerges along most major rivers in the study area: upstream reaches tend to exhibit higher BFI values, while downstream reaches display lower values. This pattern is presumed to result from greater groundwater-surface water interactions at stream headwaters, influenced by spring outlets, and the dilution of base flow as water moves downstream. This trend is particularly evident along the Gila River, Verde River, and Little Colorado River in Arizona.

#### Trends in Base Flow & BFI {#sec-bfi-trends}

Trends in BFI over the period of record for each streamgage in this analysis are illustrated in @fig-instrumented-trend and @tbl-trends. Base flow and BFI trends were analyzed across all instrumented sites over their respective periods of record using the Mann–Kendall test. Statistically significant trends were observed in both metrics, with a 72.20% coincidence rate between significant base flow and BFI trends, indicating a strong dependence of BFI on base-flow dynamics.

@fig-instrumented-trend illustrates the spatial variation in BFI trends across the study area. Statistically significant decreasing trends are observed at 16.1% of sites, while increasing trends are found at 8.8% of sites, with no clear regional patterns for either. In the Basin and Range physiographic region, 9% of sites show increasing trends, while 16.7% exhibit decreasing trends. In the Colorado Plateau region, increasing trends occur at 6.1% of sites, and decreasing trends are observed at 14.3% of sites.

![Trends in BFI over full period of record for instrumented sites used in this study. Red upward (blue downward) arrows indicate an increasing (decreasing) trend at a significance level of 5%. White circles represent sites with no statistically significant trends.](images/BFITrends_20241203.png){#fig-instrumented-trend}

#### Classification Trends

Classifications presented in @tbl-trends were determined based on precipitation regime, physiographic region, climate, and slope. The dominant precipitation regime (monsoon vs. snowmelt) was identified by analyzing streamflow hydrographs for each station, focusing on peak flow periods during the monsoon season (July–September) and the snowmelt season (March–June). Physiographic region was assigned based on which region the streamgage is located. Climate classifications were defined as warm (above the long-term median temperature of Arizona), cool (below the long-term median temperature), wet (above the long-term median precipitation), and dry (below the long-term median precipitation). Slope was categorized as high (above the median slope) and low (below the median slope).

Statistically significant decreasing trends in BFI were more common than increasing trends across all site classifications (@tbl-trends). While decreasing trends dominate, both increasing and decreasing trends are observed within each classification. Monsoon-dominated regions exhibit a higher proportion of significant negative trends (24.1%) compared to snowmelt-dominated regions (10.2%), suggesting that monsoon-dominated systems are more consistently correlated with declining base flow. Among climate classifications, warm-dry climates have the highest proportion of negative trends (20.0%), followed by warm-wet climates (19.4%), indicating that regions with higher temperatures are more prone to base flow declines. Low-slope regions show a greater prevalence of negative trends (20.4%) compared to high-slope regions (11.8%). This suggests that flatter areas may be more susceptible to base-flow reductions, potentially due to differences in hydrologic connectivity and recharge dynamics.

| **Classification Group** | **n** | **n_pos** | **n_neg** | **perc_pos** | **perc_neg** |
|------------|------------|------------|------------|------------|------------|
| Precipitation - Monsoon Dominated | 87 | 8 | 21 | 0.092 | 0.241 |
| Precipitation - Snowmelt Dominated | 118 | 9 | 12 | 0.076 | 0.102 |
| Physiographic Region - Basin and Range | 156 | 14 | 26 | 0.090 | 0.167 |
| Physiographic Region - Colorado Plateau | 49 | 3 | 7 | 0.061 | 0.143 |
| Climate - Warm-Wet | 31 | 2 | 6 | 0.065 | 0.194 |
| Climate - Warm-Dry | 55 | 6 | 11 | 0.109 | 0.200 |
| Climate - Cool-Wet | 74 | 4 | 9 | 0.054 | 0.122 |
| Climate - Cool-Dry | 45 | 5 | 7 | 0.111 | 0.156 |
| Slope - High | 102 | 10 | 12 | 0.098 | 0.118 |
| Slope - Low | 103 | 7 | 21 | 0.068 | 0.204 |

: Comparison of trends for BFI for all sites split by various classifications. Only sites with a significant ($\rho \le 0.05$) trend are included here as established by a Mann-Kendall test for monotonic trends across the full period of record. n is the number of sites, n_pos (n_neg) is the number of sites with positive (negative) trends, perc_pos (perc_neg) is the percentage of n with a positive (negative) trend. {#tbl-trends}

#### Coincident Climate Trends

Given the variations in the period of record across the instrumented network (see Figure S2), we analyzed trends in base flow and BFI in relation to coincident trends in climate variables. Trends are classified as coincident when the direction of the climate variable trend (positive or negative) aligns with the trend observed in base flow or BFI (@tbl-coincident-trends). This analysis includes both significant and non-significant trends, which is appropriate where the influence of complex, interconnected processes may not always manifest as statistically significant patterns over limited observational periods [@ficklin2016].

|           | **Climate Variable** | **Coincidence Percentage (%)** |
|-----------|:--------------------:|:------------------------------:|
| BFI       |        ET~O~         |             44.88              |
|           |    Precipitation     |             53.17              |
|           |     Temperature      |             47.32              |
| Base Flow |        ET~O~         |             55.12              |
|           |    Precipitation     |             64.88              |
|           |     Temperature      |             39.02              |

: Coincident trends of climate variables (ET~O~, precipitation, temperature) with base flow and BFI trends. {#tbl-coincident-trends}

The analysis shows that base flow and BFI trends most frequently align with precipitation trends (64.88% and 53.17%, respectively), emphasizing precipitation as the primary driver of local groundwater recharge and discharge. Coincidence with ET~O~ (reference evapotranspiration) trends (55.12% for base flow and 44.88% for BFI) suggests that evapotranspiration also plays a significant role, particularly in arid regions where it can reduce recharge or base flow during dry periods. In contrast, temperature trends show lower percentages of coincidence, often opposing base flow and BFI trends. Specifically, positive (negative) temperature trends are frequently associated with negative (positive) base-flow trends (60.98%) and BFI trends (52.68%). These results highlight the complex interplay between climatic variables and hydrological processes. While precipitation exerts the strongest influence on base flow and BFI, evapotranspiration and temperature add further variability in specific environmental contexts.

## Discussion
### Model Performance and Implications

The regional model demonstrated strong predictive capability (overall R²=0.764) but showed systematic underprediction (negative pbias). The systematic bias suggests potential limitations in the model or underlying predictor data, particularly in monsoon-dominated and warm-dry climates. The performance differences across climatic and physiographic classifications highlight the need for targeted improvements, such as integrating additional predictors or refining the representation of regional hydrological processes.

### Interpretation of Predictor Variables

Elevation emerged as a critical predictor, reinforcing its effectiveness as a proxy for climate conditions influencing groundwater recharge. The mixed influences of Soil Type A may indicate complexities in how infiltration processes are represented in the model or variability in hydrologic conditions within similar soil classes. Future research could benefit from explicit inclusion of temporal variables such as snowpack duration, soil moisture storage, and finer-scale land cover attributes.

### Spatial Patterns of BFI

The spatial distribution of predicted and observed BFI reflects known hydrogeological and climatic conditions. High BFI values correlate strongly with spring-fed streams and physiographic transitions, while low values align with arid regions. This spatial consistency validates the model’s utility for regional water resource assessments and highlights areas vulnerable to declining baseflows.

### Trends and Climate Implications

The prevalence of negative BFI trends, especially in warm-dry and low-slope regions, suggests that climate warming and reduced precipitation could disproportionately impact baseflow. The relatively weaker alignment of temperature trends with BFI suggests that temperature impacts are mediated through complex processes, potentially offset by concurrent precipitation or land-use changes. This complexity emphasizes the importance of integrated climate and land-management strategies to sustain groundwater-dependent ecosystems.
